---
title: "Lab 2 - Data wrangling"
author: "Quincy Vogelsang"
date: "February 11, 2026"
format: 
  html: default # for speed
  pdf: default # for final output
---

```{r}
#| eval: true
#| message: false

library(tidyverse)
```

# Questions

## Part 1

```{r}
?midwest

glimpse(midwest)
```

### Question 1

```{r}
midwest |>
  group_by(state) |>
  summarize(counties = n()) |>
  arrange(desc(counties))
  
```

The highest number of counties can be found in IL with 102 counties. The lowest number of counties can be found in WI with 72 counties.

\newpage

### Question 2

```{r}
midwest |>
  group_by(county) |>
  summarize(n_states = n()) |>
  filter(n_states == 5)
```

\newpage

### Question 3

```{r}

midwest |>
  filter(popdensity > 25000) |>
  arrange(desc(popdensity)) |>
  select(county, state, popdensity, poptotal, area)

```

```{r}

midwest |>
  filter(popdensity == max(popdensity)) |>
  select(county, state, popdensity, poptotal, area)
```

\newpage

### Question 4

```{r}

midwest |>
  summarize(
    typical_density = median(popdensity),
    q1 = quantile(popdensity, 0.25),
    q3 = quantile(popdensity, 0.75)
  )
```

The distribution of population density of counties is unimodal and extremely right-skewed. A typical Midwestern county has population density of 1156.208 people per unit area. The middle 50% of the counties have population densities between 622.4074 to 2330 people per unit area.

\newpage

### Question 5

```{r}
midwest <- midwest |>
  mutate(metro = if_else(inmetro == 1, "Yes", "No"))

midwest |>
  group_by(state) |>
  summarize(perc_urban = mean(metro == "Yes"))
```

\newpage

### Question 6

```{r}
midwest |>
  filter(percbelowpoverty > 40, percollege < 10) |>
  select(county, state, percbelowpoverty, percollege)
```

```{r}
midwest |>
  filter(percollege > 40) |>
  select(county, state, percbelowpoverty, percollege)
```

```{r}
midwest |>
  filter(percbelowpoverty > 40 & percollege < 10 | percollege > 40) |>
  select(county, state, percbelowpoverty, percollege)
```

```{r}
midwest |> 
  mutate(potential_outlier = if_else(
    percbelowpoverty > 40 & percollege < 10 | percollege > 40, "Yes", "No"
  )) |>
  select(county, state, percbelowpoverty, percollege, potential_outlier) |>
  arrange(potential_outlier)
```

```{r}
midwest |> 
  mutate(potential_outlier = if_else(
    percbelowpoverty > 40 & percollege < 10 | percollege > 40, "Yes", "No"
  )) |>
  ggplot(aes(x = percollege, y = percbelowpoverty, colour = potential_outlier, shape = state)) +
  geom_point() +
  labs(
    x = "% college educated",
    y = "% below poverty",
    title = "Percentage with a college degree versus poverty rate",
    subtitle = "by State in the Midwest"
  )
```

\newpage

### Question 7

```{r}
state_population <- midwest |>
  group_by(state) |>
  summarise(state_totalpop = sum(poptotal)) |>
  arrange(desc(state_totalpop))

state_population
```

```{r}
state_proportions <- state_population |>
  mutate(prop_population = state_totalpop/sum(state_totalpop))|>
  arrange(desc(prop_population))

state_proportions
```

C. IL is the most populous state and 27.2% of the midwest population lives there. The least populous state is WI, with 11.6% of the midwest population living there.

\newpage

### Question 8

\newpage

## Part 2

### Question 9
